SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-09-28 scipy-1.0/paper.tex
Minor language edits to background section

SciPy recently released its first stable version: 1.0.  It signals to
the user community that the package's API has now reached a level of
maturity and stability that warrants trusting it inside research
pipelines.  But the version numbering belies the history of a package
that has already seen extensive adoption in research and industry.

Upon historic introspection, its arrival at this point is surprising,
and somewhat anomalous.  When first started, the library had very
little funding, and was written mainly by graduate students---many of
them without an official computer science education, or even the
blessing of their advisor.  To even imagine that a small group of
``rogue'' student programmers could upend the already well-established
research software ecosystem---backed by millions in funding, and
thousands of highly qualified engineers---was preposterous.

Yet, the philosophical motivations behind a fully open toolstack, and
an excited, friendly community with a singular focus, seemed to have
proven decisive in the long run.  It led not only to the library
described in this paper, but also to an entire ecosystem of related
packages, and a variety of social activities around them.  The
packages in the ecosystem share high standards of implementation,
documentation, and testing, and a culture eager to learn and adopt
better practices---both for community management and software
development.

Despite its success, SciPy still faces challenges: a low ``bus
factor'' (the minimum number of contributors that needs to leave to
cause serious difficult) being perhaps the most serious.

@@

Python is an interactive computer programming language, designed by
Guido van Rossum in the late 80s, with a heavy emphasis on readability
and rapid prototyping.
% https://legacy.python.org/search/hypermail/python-1992/0000.html
%
While it was a general purpose programming language with a ``batteries
included'' standard library, it had no special support for scientific
data structures or algorithms, unlike many of the other established computation
platforms of the time. Yet, scientists soon discovered the language's
virtues, such as its ability to wrap C and Fortran libraries and to
then drive those libraries interactively.
Scientists could thereby gain access to a wide variety of existing
computational libraries, without concerning themselves with low-level
programming concepts such as memory management.

In 1995, Jim Hugunin, a graduate student from MIT, wrote in the first
message on the Python Matrix Special Interest Group (Matrix-SIG)
mailing list\footnote{https://mail.python.org/pipermail/matrix-sig/1995-August/000002.html}:
\begin{quote}
There seems to be a fair amount of interest in the Python community
concerning the addition of numeric operations to Python.  My own desire is
to have as large a library of matrix based functions available as possible
(linear algebra, eigenfunctions, signal processing, statistics, etc.).  In
order to ensure that all of these libraries interoperate, there needs to
be agreement on a basic matrix object that can be used to represent arrays
of numbers.
\end{quote}
Over the next several months, conversations on that mailing
list by, among others, Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, and Guido van Rossum, led to the creation of an array object
with arbitrarily high dimensions.  Jim Hugunin explained the utility
of Python for numerical computation\footnote{http://hugunin.net/papers/hugunin95numpy.html}:
\begin{quote}
I've used almost all of the available numerical languages at one time
or another over the past 8 years. One thing I've noticed is that over
time, the designers of these languages are steadily adding more of the
features that one would expect to find in a general-purpose
programming language.
\end{quote}
This is still a distinguishing feature of Python for science, and one of the
reasons why it has been so successful in the realm of data science: instead of
adding general features to a language designed for numerical and scientific
computing, here scientific features are added to a general purpose language
instead.  This broadens the scope of problems that can be addressed easily, the
sources of data that are readily accessible, and increases the size of the
community that develops general purpose tooling.

The availability of a standard numerical array data structure in
Python led to a sudden and rapidly growing number of scientific
packages becoming available for solving various common numeric
problems.
A number of these packages were written by graduate students and
postdoctoral researchers to solve the very practical research problems
that they faced on a daily basis.  While they had access to and were
familiar with specialized (and often commercial) systems, many found
it easier to implement the domain-specific functionality they needed
in a general purpose programming language.  And given Python's innate
ability to function as a systems language, controlling specialized or
custom-built hardware was feasible.

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack.  Many of the first efforts were driven by graduate
students and post-doctoral researchers, but soon attracted the
attention of senior researchers.
% Paul Dubois, CDAT
% https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html
Paul Dubois at Lawrence Livermore National Laboratory took over the
maintenance of Numeric and funded the writing of its
manual\footnote{http://www.numpy.org/\_downloads/numeric-manual.pdf}.
%in the Program for Climate Model Diagnosis and Intercomparison at
%Lawrence Livermore National Laboratory built their Climate Data
%Analysis Tools (``an open-source set of Python modules and tools for
%accessing, manipulating, and plotting scientific data in general and
%climate data sets in
%particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
%on Numeric, the newly established Python array object.  Paul later
%became the maintainer for Numeric, and helped fund the writing of the
%first manual\footnote{https://conference.scipy.org/scipy2010/slides/travis_oliphant_keynote.pdf}
In 1998, the Space Telescope Science Institute, in charge of science
operations of the Hubble Space Telescope, decided to replace their
custom scripting language with Python\footnote{https://conference.scipy.org/scipy2011/slides/greenfield\_keynote\_astronomy.pdf}.

>>>>

SciPy recently released its first stable version: 1.0.  It signals to
the user community that the library's API has now reached a level of
maturity and stability that warrants trusting it inside research
pipelines.  But the version numbering belies the history of a package
that has already seen extensive adoption in research and industry.

Upon historic introspection, its arrival at this point is surprising,
and somewhat anomalous.  When first started, the library had very
little funding, and was written mainly by graduate students---many of
them without a computer science education, and often without the
blessing of their advisors.  To even imagine that a small group of
``rogue'' student programmers could upend the already well-established
ecosystem of research software---backed by millions in funding, and
thousands of highly qualified engineers---was preposterous.

Yet, the philosophical motivations behind a fully open toolstack, and
an excited, friendly community with a singular focus, seemed to have
proven decisive in the long run.  It led not only to the library
described in this paper, but also to an entire ecosystem of related
packages, and a variety of social activities centered around them.  The
packages in the SciPy ecosystem share high standards of implementation,
documentation, and testing, and a culture eager to learn and adopt
better practices---both for community management and software
development.

Despite its success, SciPy still faces challenges: a low ``bus
factor''---the minimum number of contributors that needs to leave to
cause serious difficulty---being perhaps the most serious.

@@

Python is an interactive computer programming language, designed by
Guido van Rossum in the late 80s, with a heavy emphasis on readability
and rapid prototyping.
% https://legacy.python.org/search/hypermail/python-1992/0000.html
%
A general purpose programming language with a ``batteries included''
standard library, it had no special support for scientific data
structures or algorithms, unlike many of the other established
computation platforms of the time. Yet, scientists soon discovered the
language's virtues, such as its ability to wrap C and Fortran
libraries and to then drive those libraries interactively.  Scientists
could thereby gain access to a wide variety of existing computational
libraries, without concerning themselves with low-level programming
concepts such as memory management.

In 1995, Jim Hugunin, a graduate student from MIT, wrote in the first
message on the Python Matrix Special Interest Group (Matrix-SIG)
mailing list\footnote{https://mail.python.org/pipermail/matrix-sig/1995-August/000002.html}:
\begin{quote}
There seems to be a fair amount of interest in the Python community
concerning the addition of numeric operations to Python.  My own desire is
to have as large a library of matrix based functions available as possible
(linear algebra, eigenfunctions, signal processing, statistics, etc.).  In
order to ensure that all of these libraries interoperate, there needs to
be agreement on a basic matrix object that can be used to represent arrays
of numbers.
\end{quote}
Over the next several months, conversations on that mailing
list by, among others, Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, and Guido van Rossum, led to the creation of an array object
that supported a high number of dimensions.  Jim Hugunin explained the utility
of Python for numerical computation\footnote{http://hugunin.net/papers/hugunin95numpy.html}:
\begin{quote}
I've used almost all of the available numerical languages at one time
or another over the past 8 years. One thing I've noticed is that over
time, the designers of these languages are steadily adding more of the
features that one would expect to find in a general-purpose
programming language.
\end{quote}
This is still a distinguishing feature of Python for science, and one of the
reasons why it has been so successful in the realm of data science: instead of
adding general features to a language designed for numerical and scientific
computing, here scientific features are added to a general purpose language
instead.  This broadens the scope of problems that can be addressed easily, the
sources of data that are readily accessible, and increases the size of the
community that develops general purpose tooling.

The availability of a standard numerical array data structure in
Python led to a sudden and rapidly growing number of scientific
packages becoming available for solving various common numeric
problems.
A number of these packages were written by graduate students and
postdoctoral researchers to solve the very practical research problems
that they faced on a daily basis.  While they had access to and were
familiar with specialized (often commercial) systems, many found
it easier to implement the domain-specific functionality they needed
in a general purpose programming language.  And given Python's innate
ability to function as a systems language, controlling specialized or
custom-built hardware was feasible.

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack. Many of the first efforts were driven by graduate
students and post-doctoral researchers, but soon attracted the
attention of more senior researchers.
% Paul Dubois, CDAT
% https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html
For example, Paul Dubois at Lawrence Livermore National Laboratory took over the
maintenance of Numeric and funded the writing of its
manual\footnote{http://www.numpy.org/\_downloads/numeric-manual.pdf}, and
%in the Program for Climate Model Diagnosis and Intercomparison at
%Lawrence Livermore National Laboratory built their Climate Data
%Analysis Tools (``an open-source set of Python modules and tools for
%accessing, manipulating, and plotting scientific data in general and
%climate data sets in
%particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
%on Numeric, the newly established Python array object.  Paul later
%became the maintainer for Numeric, and helped fund the writing of the
%first manual\footnote{https://conference.scipy.org/scipy2010/slides/travis_oliphant_keynote.pdf}
in 1998 the Space Telescope Science Institute, in charge of science
operations of the Hubble Space Telescope, decided to replace their
custom scripting language with Python\footnote{https://conference.scipy.org/scipy2011/slides/greenfield\_keynote\_astronomy.pdf}.
